---
title: "Data Wrangling"
output: html_document
---



# Libraries


```{r libraries}
library(tidyverse)
library(haven)
library(readxl)
library(janitor)

options(tibble.print_min = 5)

```

# Importing Data


```{r importing_data}
litters_df = read_csv(file = "./data/FAS_litters.csv") # Relative Path

# pups_df = read_csv(file = "~\Desktop\Fall 2024\Data Science\Week 3 - Data #Wrangling\p8105_DataWrangling\data\FAS_litters.csv") # Absolute path

```

# Cleaning data

```{r}
names(litters_df)

# Janitor - will convert to snake case
litters_df = janitor::clean_names(litters_df) 
names(litters_df)
```

# Skim data
```{r}

skimr::skim(litters_df)

# Using read to read in a file:  the call below will skip the first 50 lines of data and not assume the first row are variable names
```

# Using read_csv to read in a file

```{r}

litters_df = 
    read_csv(file = "./data/FAS_litters.csv",
    skip = 10, col_names = FALSE)

```

# Parsing columns

col_types = cols(...): This specifies the data types for each column:
Group and Litter Number are set as character (text) columns
GD0 weight and GD18 weight are set as double (decimal number) columns
GD of Birth, Pups born alive, Pups dead @ birth, and Pups survive are set as integer columns

```{r}
litters_df = 
    read_csv(file = "./data/FAS_litters.csv",
        na = c(".", "NA", ""),
    col_types = cols(
      Group = col_character(),
      `Litter Number` = col_character(),
      `GD0 weight` = col_double(),
      `GD18 weight` = col_double(),
      `GD of Birth` = col_integer(),
      `Pups born alive` = col_integer(),
      `Pups dead @ birth` = col_integer(),
      `Pups survive` = col_integer()
    )
  )

tail(litters_df)
```

# There's a compact way to assign variable types


col_types = "fddddd": This is a compact way to specify column types:

f: The first column is set as a factor (categorical variable)
d: The next five columns are set as doubles (decimal numbers)

```{r}
pups_df = 
    read_csv("./data/FAS_pups.csv",
        na = c(".", "NA"), col_types = "fddddd")

skimr::skim(pups_df)

```



# Importing other file formats: using readxl and haven

Non-CSV plain text files (e.g. tab delimited files) can be handled using read_delim(). This is very similar to read_csv, but you have to specify a delimiter.

n_max = 20: This argument limits the number of rows read from the Excel file to a maximum of 20.

```{r}
library(readxl)

mlb11_df = read_excel("data/mlb11.xlsx", n_max = 20)

head(mlb11_df, 5)


library(haven)

pulse_df = read_sas("./data/public_pulse_data.sas7bdat")
# You can read in data that isn’t coming as a flat file (think data rectangle), but it’s beyond our current scope.

head(pulse_df, 5)
```


# Comparison with BaseR

The base R versions tend to be slower (very noticeably for large datasets), and the default options can make less sense for modern datasets. Meanwhile, the readr functions export tibbles, and keep characters as characters (instead of converting to factors …).

Key differences you might observe:

1. Data types: read_csv() often does a better job at guessing column types, while read.csv() might default to factors for character columns.

2. Speed: read_csv() is generally faster, especially for larger files.

3. Column names: read_csv() might handle spaces and special characters in column names differently than read.csv().

4. Missing values: read_csv() has more flexible options for handling missing values.

5. The $S accessor: If "S" is not a column name but the start of a longer column name, read.csv() might allow partial matching, while read_csv() typically does not.


NOTE: Never use read.csv (ie BASER) !!

```{r}

library(tidyverse)

pups_base = read.csv("./data/FAS_pups.csv") # This uses base R's read.csv() function to read the CSV file and store it in pups_base.
pups_readr = read_csv("./data/FAS_pups.csv") # This uses the readr package's read_csv() function to read the same CSV file and store it in pups_readr.


View(pups_base)
View(pups_readr)

pups_base
pups_readr

pups_base$S
pups_readr$S

```

## Data Science Data Wrangling Continued

```{r}

pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df

pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_data



pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")



## Cleaning/prepping data altogether

pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) 

print(pulse_df, n = 12)

```


## Learning Assessment:
In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?


```{r LearningAssessment}

cleaned_litters_df <- litters_df %>%
  select(litter_number, gd0_weight, gd18_weight) %>%
  pivot_longer(
    cols = c(gd0_weight, gd18_weight),
    names_to = "gd",
    values_to = "weight"
  ) %>%
  mutate(
    gd = as.numeric(str_remove(gd, "gd") %>% str_remove("_weight"))
  )

print(cleaned_litters_df)





```

##UNTIDY DATA

```{r}


analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
  )

analysis_result
## # A tibble: 4 × 3
##   group     time   mean
##   <chr>     <chr> <dbl>
## 1 treatment pre     4  
## 2 treatment post    8  
## 3 placebo   pre     3.5
## 4 placebo   post    4


analysis_result %>%
  pivot_wider(
    names_from = "time", 
    values_from = "mean"
  ) %>%
  knitr::kable()

## # A tibble: 2 × 3
##   group       pre  post
##   <chr>     <dbl> <dbl>
## 1 treatment   4       8
## 2 placebo     3.5     4



```




## BINDING ROWS

```{r binding_rows}

# read each table and do some cleaning

fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")


```