---
title: "Tidy Data"
author: "NSK"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Data Science Data Wrangling Continued

```{r}

pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df

pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_data



pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")



## Cleaning/prepping data altogether

pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) 

print(pulse_df, n = 12)

```


## Learning Assessment:
In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?


```{r LearningAssessment}

cleaned_litters_df <- litters_df %>%
  select(litter_number, gd0_weight, gd18_weight) %>%
  pivot_longer(
    cols = c(gd0_weight, gd18_weight),
    names_to = "gd",
    values_to = "weight"
  ) %>%
  mutate(
    gd = as.numeric(str_remove(gd, "gd") %>% str_remove("_weight"))
  )

print(cleaned_litters_df)





```

##UNTIDY DATA

```{r}


analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
  )

analysis_result
## # A tibble: 4 × 3
##   group     time   mean
##   <chr>     <chr> <dbl>
## 1 treatment pre     4  
## 2 treatment post    8  
## 3 placebo   pre     3.5
## 4 placebo   post    4


analysis_result %>%
  pivot_wider(
    names_from = "time", 
    values_from = "mean"
  ) %>%
  knitr::kable()

## # A tibble: 2 × 3
##   group       pre  post
##   <chr>     <dbl> <dbl>
## 1 treatment   4       8
## 2 placebo     3.5     4



```




## BINDING ROWS

```{r binding_rows}

# read each table and do some cleaning

fellowship_ring = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")


```